---
title: "1.6 Data exploration Sphagnum cover"
author: "MarteF"
format: html
editor: visual
---

## Packages needed

```{r Packages needed, output=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(lubridate)

```

## Uploading and cleaning the data

```{r Upload datasets, output=FALSE, warning=FALSE}

reveg_var <- readr::read_delim('../data/reveg_var.csv', 
                       delim = ',',
                       locale = locale('se', encoding = 'ISO8859-1'),
                       col_names = TRUE) 

reveg_var <- reveg_var %>% 
  select( -1) %>%
  mutate_at(c('year', 'month', 'Location', 'Sublocation', 'Block', 'Treatment_ID', 'Treatment', 'Name', 'Station_ID'), as.factor) %>%
  select('Date','year','month', 'Name','Location','Sublocation', 'Block', 'Treatment', 'Treatment_ID', 'Station_ID', everything())
```

## Visual inspection of data

### Sphagnum cover

```{r}
reveg_var %>%
  filter(is.na(Sphagnum))
```

There seem to be one record of nutrients that isn't connected to a treatment plot, resulting in an NA within Sphagnum cover ...

Overall Sphagnum cover changes:

```{r, warning=FALSE}
plot_SP <- reveg_var %>%
  filter(!is.na(Sphagnum)) %>%  # Exclude NA values in the Sphagnum column
  ggplot() +   
  geom_boxplot()+   
  aes(x=Treatment, y=Sphagnum, color=as.factor(t_year)) +   
   ylab("Sphagnum cover") +
  labs(color = "Growth seasons since treatment") +
  theme_bw()
plot_SP
```

The overall trends is quite clear. Sphagnum treatment increases Sphagnum cover substantially compared to mulch only and controls, even after 2 growth seasons, but especially after 3. The results looks almost similar to the references after 3 growth seasons.

```{r}
reveg_var %>%
  filter(!is.na(Sphagnum)) %>%  # Exclude NA values in the Sphagnum column
  ggplot(aes(x = Sphagnum)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Sphagnum Values",
       x = "Sphagnum",
       y = "Frequency") +
  facet_wrap(~ t_year)
```

Overall Sphagnum cover per location:

```{r, warning=FALSE}
plot_SP_Location <- reveg_var %>%
  filter(!is.na(Sphagnum)) %>%  # Exclude NA values in the Sphagnum column
  ggplot() +   
  geom_boxplot()+   
  aes(x=Treatment, y=Sphagnum, color=as.factor(t_year)) +
  ylab("Sphagnum cover") +
  labs(color = "Growth seasons since treatment") +
  facet_wrap(Location~.) +
  theme_bw()
plot_SP_Location +  theme(legend.position="bottom")
```

There is some differences between sites. The Sphagnum treatment is clearly not as successful at the location VSM (Vestersetermyra).

```{r}
reveg_var%>%
  group_by(Location, t_year, Treatment) %>%
  summarise(median=median(Sphagnum), 
            mean=mean(Sphagnum),
            Sph0 = sum(Sphagnum==0),
            Sph0.1= sum(Sphagnum==0.1),
            Sph1 = sum(Sphagnum==1),
            max=max(Sphagnum), total_N = n()) 
```

```{r}

se <- function(x) {
  sd(x) / sqrt(length(x))
}

reveg_var%>%
  filter(t_year==3) %>%
  group_by(Location, Treatment) %>%
  summarise(median=median(Sphagnum), 
            mean=mean(Sphagnum),
            se = se(Sphagnum),
            max=max(Sphagnum), total_N = n()) 
```

```{r}
reveg_var%>%
  group_by(t_year, Treatment) %>%
  summarise(median=median(Sphagnum), 
            mean=mean(Sphagnum),
            se = se(Sphagnum),
            max=max(Sphagnum), total_N = n()) 
```

```{r}
reveg_var%>%
  dplyr::select(Treatment_ID, Treatment, t_year, Location, Sublocation, Block, Sphagnum, Change_Year, Change_Total) %>%
  filter(Sphagnum==0 & t_year==3 & Treatment=="S")
```

There's 1 plot from the Sphagnum treatment that had 0 Sphagnum after 3 growth seasons.

Checking which plots had a negative growth of Sphagnum:

```{r}
reveg_var%>%
  filter(!is.na(Change_Total)) %>%
    dplyr::select(Treatment_ID, t_year, Location, Sublocation, Block, Sphagnum, Change_Year, Change_Total) %>%
  filter(Change_Total<=0)%>%
  arrange(Change_Total, Treatment_ID)

```

```{r}
reveg_var %>%      
  filter(Change_Year<0 | Change_Total<0) %>%      
  dplyr::select(c(t_year, Location, Sublocation, Block, Treatment, Treatment_ID, Sphagnum, Change_Year, Change_Total))%>%      
  arrange(Change_Total)
```

23 plots had zero or negative development of Sphagnum in total throughout the period, but only 4 had a negative development in total (and then really low - only 0.1 and 0.9).

Several plots stagnated or dropped a little from 2nd to 3rd year, of these are 1 Sphagnum plot from VSM.

### Summary of Sphagnum cover per sublocation

```{r}
reveg_var%>%
  group_by(Location, Sublocation, t_year, Treatment) %>%
  summarise(median=median(Sphagnum), 
            mean=mean(Sphagnum),
           se =se(Sphagnum),
           sd = sd(Sphagnum),
            max=max(Sphagnum), total_N = n()) 
```

```{r}
reveg_var%>%
  group_by(Location, t_year, Treatment) %>%
  summarise(median=median(Sphagnum), 
            mean=mean(Sphagnum),
           se =se(Sphagnum),
           sd = sd(Sphagnum),
            max=max(Sphagnum))
```
