---
title: "1.7 Data exploration vegetation cover"
author: "MarteF"
format: html
editor: visual
---

## Packages needed

```{r}
library(tidyverse)
library(readr)
library(lubridate)
library(forcats)
```

## Uploading and cleaning the data

```{r}
reveg_var <- readr::read_delim('../data/reveg_var.csv', 
                       delim = ',',
                       locale = locale('se', encoding = 'ISO8859-1'),
                       col_names = TRUE)

reveg_var <- reveg_var %>% 
  select( -1) %>%
  mutate_at(c('year', 'month', 'Location', 'Sublocation', 'Block', 'Treatment_ID', 'Treatment', 'Name', 'Station_ID'), as.factor) %>%
  select('Date','year','month', 'Name','Location','Sublocation', 'Block', 'Treatment', 'Treatment_ID', 'Station_ID', everything()) %>%
  rename(WT_consecutive_days_below = consecutive_days_count, WT_total_days_below=total_days, WT_mean=gs_mean, WT_max=gs_max, WT_min=gs_min)

```

I've seen that there is a need for some cleaning within the plant functional groups. Some smaller and very detailed groups should be combined, like Rhynchospora should join Carex, and Oxycoccus and Drosera should join Herbs. Some have not the correct names. This doesn't affect the former calculations of total vegetation cover or ground/field layer.

```{r}
#Renaming and combining several of the field groups into more coherent functional groups:

reveg_var <- reveg_var %>%
  rename(Liverworts = Livermosses, Rushes = Sedges, Ericoids = Heathers, Eriophorum = Eriopherum) %>% #rename misnamed groups
  mutate(Shrubs_Trees = Trees + Bushes + Bet_nan,   # combine Trees, Bushes, and Bet_nan into Shrubs_Trees
         Carex = Carex + Rhynchospora,       # join Rhyncospora into Carex
         Herbs = Herbs + Oxycoccus + Drosera, # join Oxycoccus and Drosera into Herbs
         Other_bryophytes = Other_mosses + Racomitrium)    %>%       
 select(-Trees, -Bushes, -Bet_nan, -Rhynchospora, -Oxycoccus, -Drosera, -Other_mosses, -Racomitrium)  %>% # remove the old columns
  relocate(Shrubs_Trees, .after = Seedlings)  
```

## Other vegetation cover

Summarising all the other interesting cover groups (Mulch, Bare_peat, Other_veg, Total_veg)

```{r}
reveg_var%>%
  dplyr::select(c(Location, Sublocation, Block, Treatment, Treatment_ID, Sphagnum, Mulch, Bare_peat, t_year, Other_veg, Total_veg, Field, Veg_Ground, Tot_Ground))%>%
  group_by(Treatment, t_year) %>%
  summarise(Sph_mean=mean(Sphagnum),
            Sph_sd= sd(Sphagnum),
            Sph_max=max(Sphagnum),
            Mulch_mean = mean(Mulch),
            Mulch_sd = sd(Mulch),
            Bare_peat_mean = mean(Bare_peat),
            Bare_peat_sd = sd(Bare_peat),
            Other_veg_mean = mean(Other_veg),
            Other_veg_sd = sd(Other_veg),
            Other_veg_max = max(Other_veg),
            Field_mean = mean(Field),
            Field_sd = sd(Field),
            Field_max = max(Field),
            Veg_Ground_mean = mean(Veg_Ground),
            Veg_Ground_sd = sd(Veg_Ground),
            Veg_Ground_max = max(Veg_Ground),
            Tot_Ground_mean = mean(Tot_Ground),
            Tot_Ground_sd = sd(Tot_Ground),
            Tot_Ground_max = max(Tot_Ground),
             total_N = n()) 
```

It seems like a few of the plots have a very high total vegetation cover. The total sum of vegetation cover comes to slightly more than 100% makes sense, since the evaluation of cover in the field layer and the ground layer may overlap, but it shouldn't become very much more, I think.

I need to take a closer look at those with total vegetation cover of more than 120%.

```{r}
reveg_var%>%
  dplyr::select(c(Location, t_year, Sublocation, Block, Treatment, Treatment_ID, Sphagnum, Other_bryophytes, Liverworts, Lichen, Polytrichum, Trichophorum, Eriophorum, Ericoids, Narthecium, Other_veg, Veg_Ground, Tot_Ground, Field, Total_veg))%>%
 filter(Total_veg>120)
```

There are 21 plots that have total vegetation cover of \>120%.

Of these, 12 exceed 130% vegetation cover.

13 of the plots have \>100% ground layer cover, but only 5 have \>105%. These are the ones I should focus on, I think. Two of these actually have a vegetation ground cover \>100%, but just barely.

Most plots are from the references, where the vegetation in both ground layer and field layer is extensive.

Only two plots are from the experimental plots - HM-3-2-S2 from year3, and VSM-2-1-S2 from year 3.

First has Sphagnum=85 and Eriopherum=60, totalling Total_veg=145. That might not be unreasonable, as these may overlap.

Second has Sphagnum=75 and Eriopherum=40, totaling Total_veg=121.5, which is perfectly fine.

The total ground cover also exceeds 100% in many cases, which shouldn't be possible.

Let's take a closer look at that.

```{r}
reveg_var%>%
  dplyr::select(c(Location, Sublocation, Block, Treatment, Treatment_ID, t_year, Sphagnum:Lichen, Bare_peat:Temp_water, Other_veg, Total_veg, Field, Veg_Ground, Tot_Ground))%>%
  group_by(t_year, Location) %>%
 filter(Tot_Ground>120)
```

7 plots have higher ground cover than 120%. Most of this is due to overlap with mulch, and I have gone through the photos of each and double checked and altered any cover percentages that seemed a little wrong. All ok now.

```{r}
reveg_var%>%
  dplyr::select(c(Location, Sublocation, Block, Treatment, Treatment_ID, Sphagnum, Mulch, Bare_peat, t_year, Other_veg, Total_veg, Field, Veg_Ground, Tot_Ground))%>%
  group_by(Treatment, t_year) %>%
  summarise(Other_veg_mean = mean(Other_veg),
            Other_veg_sd = sd(Other_veg),
            Other_veg_max = max(Other_veg),
            Field_mean = mean(Field),
            Field_sd = sd(Field),
            Field_max = max(Field),
            Veg_Ground_mean = mean(Veg_Ground),
            Veg_Ground_sd = sd(Veg_Ground),
            Veg_Ground_max = max(Veg_Ground),
            Tot_Ground_mean = mean(Tot_Ground),
            Tot_Ground_sd = sd(Tot_Ground),
            Tot_Ground_max = max(Tot_Ground),
            Tot_Veg_mean = mean(Total_veg),
             total_N = n()) 
```

## Exploring specific species groups

Creating long format of species functional groups:

```{r}
reveg_long <-reveg_var%>%
  pivot_longer(Sphagnum:Temp_water, names_to = 'functional_group', values_to = "cover")
```

Overview of the various species groups in the reference plots:

```{r}
plot_ref <- reveg_long %>%
  filter(Treatment=='R') %>%
  ggplot() +   
  geom_boxplot()+   
  aes(x=reorder(functional_group, desc(cover), FUN = median), y=cover) +
  ylab("Cover of the various functional groups in the reference plots") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
plot_ref +  theme(legend.position="bottom")
```

Overview of the FTGs in M in year 3

```{r}
plot_M <- reveg_long %>%
  filter(Treatment=='M' & t_year==3) %>%
  ggplot() +   
  geom_boxplot()+   
  aes(x=reorder(functional_group, desc(cover), FUN = median), y=cover) +
  ylab("Cover of the various functional groups in the reference plots") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
plot_M +  theme(legend.position="bottom")
```

Overview of FTGs in M&S in year 3:

```{r}
plot_S <- reveg_long %>%
  filter(Treatment=='S' & t_year==3) %>%
  ggplot() +   
  geom_boxplot()+   
  aes(x=reorder(functional_group, desc(cover), FUN = median), y=cover) +
  ylab("Cover of the various functional groups in the reference plots") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
plot_S +  theme(legend.position="bottom")
```

Overview of FTGs in C in year 3:

```{r}
plot_C <- reveg_long %>%
  filter(Treatment=='C' & t_year==3) %>%
  ggplot() +   
  geom_boxplot()+   
  aes(x=reorder(functional_group, desc(cover), FUN = median), y=cover) +
  ylab("Cover of the various functional groups in the reference plots") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
plot_C +  theme(legend.position="bottom")
```

Summary year 3 + references:

```{r}

mean_cover_ref <- reveg_var %>%
  filter(Treatment == 'R') %>%
  summarise(
    mean_Sphagnum = mean(Sphagnum, na.rm = TRUE),
    max_Sphagnum = max(Sphagnum, na.rm = TRUE),
    mean_Eriophorum = mean(Eriophorum, na.rm = TRUE),
    max_Eriophorum = max(Eriophorum, na.rm = TRUE),
    mean_Narthecium = mean(Narthecium, na.rm = TRUE),
    max_Narthecium = max(Narthecium, na.rm = TRUE),
    mean_Trichophorum = mean(Trichophorum, na.rm = TRUE),
    max_Trichophorum = max(Trichophorum, na.rm = TRUE),
    mean_Ericoids = mean(Ericoids, na.rm = TRUE),
    max_Ericoids = max(Ericoids, na.rm = TRUE),
    mean_Litter = mean(Litter, na.rm = TRUE),
    max_Litter = max(Litter, na.rm = TRUE),
    mean_OtherBryophytes = mean(Other_bryophytes, na.rm = TRUE),
    max_OtherBryophytes = max(Other_bryophytes, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "statistic"),
    names_sep = "_",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = statistic,
    values_from = value
  )


```

```{r}
mean_cover_S_3 <- reveg_var %>%
  filter(Treatment == 'S' & t_year==3) %>%
  summarise(
    mean_Sphagnum = mean(Sphagnum, na.rm = TRUE),
    max_Sphagnum = max(Sphagnum, na.rm = TRUE),
    mean_Eriophorum = mean(Eriophorum, na.rm = TRUE),
    max_Eriophorum = max(Eriophorum, na.rm = TRUE),
    mean_Narthecium = mean(Narthecium, na.rm = TRUE),
    max_Narthecium = max(Narthecium, na.rm = TRUE),
    mean_Trichophorum = mean(Trichophorum, na.rm = TRUE),
    max_Trichophorum = max(Trichophorum, na.rm = TRUE),
    mean_Ericoids = mean(Ericoids, na.rm = TRUE),
    max_Ericoids = max(Ericoids, na.rm = TRUE),
    mean_Litter = mean(Litter, na.rm = TRUE),
    max_Litter = max(Litter, na.rm = TRUE),
    mean_OtherBryophytes = mean(Other_bryophytes, na.rm = TRUE),
    max_OtherBryophytes = max(Other_bryophytes, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "statistic"),
    names_sep = "_",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = statistic,
    values_from = value
  )

```

```{r}
mean_cover_M_3 <- reveg_var %>%
  filter(Treatment == 'M' & t_year==3) %>%
  summarise(
    mean_Sphagnum = mean(Sphagnum, na.rm = TRUE),
    max_Sphagnum = max(Sphagnum, na.rm = TRUE),
    mean_Eriophorum = mean(Eriophorum, na.rm = TRUE),
    max_Eriophorum = max(Eriophorum, na.rm = TRUE),
    mean_Narthecium = mean(Narthecium, na.rm = TRUE),
    max_Narthecium = max(Narthecium, na.rm = TRUE),
    mean_Trichophorum = mean(Trichophorum, na.rm = TRUE),
    max_Trichophorum = max(Trichophorum, na.rm = TRUE),
    mean_Ericoids = mean(Ericoids, na.rm = TRUE),
    max_Ericoids = max(Ericoids, na.rm = TRUE),
    mean_Litter = mean(Litter, na.rm = TRUE),
    max_Litter = max(Litter, na.rm = TRUE),
    mean_OtherBryophytes = mean(Other_bryophytes, na.rm = TRUE),
    max_OtherBryophytes = max(Other_bryophytes, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "statistic"),
    names_sep = "_",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = statistic,
    values_from = value
  )

```

```{r}
mean_cover_C_3 <- reveg_var %>%
  filter(Treatment == 'C' & t_year==3) %>%
  summarise(
    mean_Sphagnum = mean(Sphagnum, na.rm = TRUE),
    max_Sphagnum = max(Sphagnum, na.rm = TRUE),
    mean_Eriophorum = mean(Eriophorum, na.rm = TRUE),
    max_Eriophorum = max(Eriophorum, na.rm = TRUE),
    mean_Narthecium = mean(Narthecium, na.rm = TRUE),
    max_Narthecium = max(Narthecium, na.rm = TRUE),
    mean_Trichophorum = mean(Trichophorum, na.rm = TRUE),
    max_Trichophorum = max(Trichophorum, na.rm = TRUE),
    mean_Ericoids = mean(Ericoids, na.rm = TRUE),
    max_Ericoids = max(Ericoids, na.rm = TRUE),
    mean_Litter = mean(Litter, na.rm = TRUE),
    max_Litter = max(Litter, na.rm = TRUE),
    mean_OtherBryophytes = mean(Other_bryophytes, na.rm = TRUE),
    max_OtherBryophytes = max(Other_bryophytes, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "statistic"),
    names_sep = "_",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = statistic,
    values_from = value
  )
```

Combine these outputs:

```{r}
# Add an identifier column to each table
mean_cover_ref <- mean_cover_ref %>%
  mutate(Treatment = "R")

mean_cover_S_3 <- mean_cover_S_3 %>%
  mutate(Treatment = "S_3")

mean_cover_M_3 <- mean_cover_M_3 %>%
  mutate(Treatment = "M_3")

mean_cover_C_3 <- mean_cover_C_3 %>%
  mutate(Treatment = "C_3")

# Combine all tables into one
combined_mean_cover <- bind_rows(mean_cover_ref, mean_cover_S_3, mean_cover_M_3, mean_cover_C_3)

# Print the combined table
print(combined_mean_cover)
```

The reference plots have a high cover of Sphagnum (about 71% mean), and a decent cover of Narthecium, Ericoids (mostly heather) and Trichophorum. There's also some litter (about 14%ish), while Eriopherum covers only ca. 3%)

```{r}
summary_data <- reveg_var %>%
  filter(Treatment == "M", t_year == 3) %>%    # Filter for Treatment "M" and t_year 3
  select(Sphagnum:Shrubs_Trees) %>%                   # Select the vegetation columns
  summarise_all(mean, na.rm = TRUE) %>%         # Summarize each vegetation group (summing their values)
  gather(Vegetation_Group, Total_Cover) %>%    # Gather the vegetation columns into key-value pairs
  arrange(desc(Total_Cover))                   # Arrange the groups from largest to smallest

# View the summary
print(summary_data)
```

```{r}
summary_data <- reveg_var %>%
  filter(Treatment == "S", t_year == 3) %>%    # Filter for Treatment "M" and t_year 3
  select(Sphagnum:Shrubs_Trees) %>%                   # Select the vegetation columns
  summarise_all(mean, na.rm = TRUE) %>%         # Summarize each vegetation group (summing their values)
  gather(Vegetation_Group, Total_Cover) %>%    # Gather the vegetation columns into key-value pairs
  arrange(desc(Total_Cover))                   # Arrange the groups from largest to smallest

# View the summary
print(summary_data)
```

The natural cover of Eriophorum is quite low, which is interesting to compare to the experimental plots, as this is a plant that increases rapidly in newly rewetted peatlands. In the experimental plots:

```{r}
plot_eri <- reveg_var %>%
  filter(!is.na(Eriophorum)) %>%  # Exclude the NA value
  ggplot() +   
  geom_boxplot()+   
  aes(x=Treatment, y=Eriophorum, color=as.factor(t_year)) +
  ylab("Eriophorum cover") +
  labs(color = "Growth seasons since treatment") +
  theme_bw()
plot_eri +  theme(legend.position="bottom")
```

```{r}
mean_cover_eri <- reveg_var %>%
  group_by(Treatment, t_year) %>%
  summarise(mean_cover = mean(Eriophorum, na.rm = TRUE),
            max_cover = max(Eriophorum, na.rm = TRUE))
print(mean_cover_eri)
```

The cover of Eriophorum is higher than in the reference plots in both C, M, and S after 3 years, if not by much. In some specific plots the cover is substantially higher than the references.

There is however a number here I need to check further. How can there be max 80% cover of Eriophorum in M in year 2, while this is only 30% in year 3? That is a drastic decrease. I have narrowed the problem down to VSM2:

```{r}
reveg_var %>%
  filter(Treatment=='M' & Station_ID=='VSM-2') %>%
  select(c(Location, Sublocation, Block, t_year, Eriophorum, Mulch, Bare_peat, Sphagnum, Polytrichum))
```

Overall vegetation cover of other plant FTGs than Sphagnum in R:

```{r}

reveg_var %>%
  filter(Treatment == 'R') %>%
  summarise(mean_cover = mean(Other_veg, na.rm = TRUE),
            max_cover = max(Other_veg, na.rm= TRUE))
```

```{r}
plot_fun3 <- reveg_long %>%
  filter(Treatment!='R') %>%
  filter(t_year==3) %>%
  ggplot(aes(x=reorder(functional_group, desc(cover), FUN = median), y=cover)) +   
  geom_boxplot()+   
  ylab("Cover of the various functional groups in the reference plots") +
  xlab("Functional plant groups (and some other cover groups)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
plot_fun3 +  theme(legend.position="bottom")
```

Most plots have still very little other vegetation than Sphagnum after 3 years.

Some plots have a very high cover of other types of mosses. Same goes for Eriophorum, Ericoids and Polytrichum.

```{r}
plot_hea <- reveg_var %>%
  filter(!is.na(Ericoids)) %>%  # Exclude the NA value
  ggplot() +   
  geom_boxplot()+   
  aes(x=Treatment, y=Ericoids, color=as.factor(t_year)) +
  ylab("Ericoids cover") +
  labs(color = "Growth seasons since treatment") +
  theme_bw()
plot_hea +  theme(legend.position="bottom")
```

There is an interesting increase in Ericoids within the Sphagnum treatments. This is also partly visible within the mulch plots. The means are still low, but several plots seem to have had a great development in heather cover within the time period. In contrast to the Eriophorum, the Ericoids cover is still not fully reaching the levels of the references after 3 years.

## Violin plot of total vegetation cover, other vegetation cover and Sphagnum

```{r}

plot_SP <- reveg_var %>%
  filter(!is.na(Sphagnum)) %>%  # Exclude NA values in the Sphagnum column
  mutate(Treatment = factor(Treatment, levels = c("C", "M", "S", "R"))) %>% # Rearrange Treatment levels
  ggplot(aes(x = Treatment, y = Sphagnum, color = as.factor(t_year), fill=as.factor(t_year))) +   
  geom_violin(scale = "width", width= 0.8, alpha=0.7) +   
  geom_point(pch = 21, size=1, position = position_jitterdodge()) +
  ylab("Sphagnum cover") +
  labs(fill= "Growth seasons since treatment") +
  theme_minimal() +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  guides(color = "none")

plot_SP

ggsave(plot_SP, filename='../figures/violinplot_SP.png', width = 10, height = 6, units = "in")
```

```{r}
plot_TV <- reveg_var %>%
  filter(!is.na(Total_veg)) %>%  # Exclude NA values in the Sphagnum column
   mutate(Treatment = factor(Treatment, levels = c("C", "M", "S", "R")),
          Treatment = fct_recode(Treatment, "M&S" = "S")) %>% 
  # Rearrange Treatment levels
  ggplot(aes(x = Treatment, y = Total_veg, color = as.factor(t_year), fill=as.factor(t_year), alpha=0.8)) +   
  geom_violin(scale = "width", width= 0.8) +  
 geom_point(pch = 21,, size=1, position = position_jitterdodge())+
  ylab("Total vegetation cover") +
  labs(fill= "Growth seasons since treatment") +
  theme_minimal() +
   scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  guides(color = "none", alpha = "none")

plot_TV
ggsave(plot_TV, filename='../figures/violinplot_TV.png')
```

```{r}
plot_OV <- reveg_var %>%
  filter(!is.na(Other_veg)) %>%  # Exclude NA values in the Sphagnum column
   mutate(Treatment = factor(Treatment, levels = c("C", "M", "S", "R")))  %>% 
  ggplot(aes(x = Treatment, y = Other_veg, color = as.factor(t_year), fill=as.factor(t_year))) +   
  geom_violin(scale = "width", width= 0.8, alpha=0.7) +   
  geom_point(pch = 21, size=1, position = position_jitterdodge())+
  ylab("Total cover of other vegetation") +
  labs(fill= "Growth seasons since treatment") +
  theme_minimal() +
   scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  guides(color = "none", alpha = "none")

plot_OV

ggsave(plot_OV, filename='../figures/violinplot_OV.png', width = 10, height = 6, units = "in")
```

## Full summary

```{r}
# Function to calculate the standard error
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

# Summarize the data
summary <- reveg_long %>%
  group_by(functional_group, t_year, Treatment) %>%
  summarise(mean = mean(cover, na.rm = TRUE),
            median = median(cover, na.rm = TRUE),
            se = se(cover),
            max = max(cover, na.rm = TRUE),
            .groups = 'drop')

# Pivot the summary data to a wide format
summary_wide <- summary %>%
  pivot_wider(names_from = t_year, 
              values_from = c(mean, median, se, max), 
              names_glue = "{.value}_year_{t_year}")

# Explicitly order the columns
summary_wide <- summary_wide %>%
  select(functional_group, Treatment,
         mean_year_0, median_year_0, se_year_0, max_year_0,
         mean_year_2, median_year_2, se_year_2, max_year_2,
         mean_year_3, median_year_3, se_year_3, max_year_3)

# Print the final summary table
print

write.csv(summary_wide, file='summary_FGTs.csv')
```

```{r}
reveg_var%>%
  dplyr::select(c(Location, Sublocation, Block, Treatment, Treatment_ID, t_year, Other_veg, Total_veg, Field, Veg_Ground, Tot_Ground))%>%
  group_by(Treatment, t_year) %>%
  summarise(Other_veg_mean = mean(Other_veg),
            Other_veg_median = median(Other_veg),
            Other_veg_se = se(Other_veg),
            Other_veg_max = max(Other_veg),
            Field_mean = mean(Field),
            Field_median = median(Field),
            Field_se = se(Field),
            Field_max = max(Field),
            Veg_Ground_mean = mean(Veg_Ground),
            Veg_Ground_median = median(Veg_Ground),
            Veg_Ground_se = se(Veg_Ground),
            Veg_Ground_max = max(Veg_Ground),
            Tot_Ground_mean = mean(Tot_Ground),
            Tot_Ground_median = median(Tot_Ground),
            Tot_Ground_se = se(Tot_Ground),
            Tot_Ground_max = max(Tot_Ground),
            Tot_Veg_mean = mean(Total_veg),
            Tot_Veg_median = median(Total_veg),
            Tot_Veg_se = se(Total_veg),
            Tot_Veg_max = max(Total_veg),
             total_N = n()) 
```

```{r}

# Define your data summarization
reveg_summary <- reveg_var %>%
  dplyr::select(c(Location, Sublocation, Block, Treatment, Treatment_ID, t_year, Other_veg, Total_veg, Field, Veg_Ground, Tot_Ground)) %>%
  group_by(Treatment, t_year) %>%
  summarise(
    Other_veg_mean = mean(Other_veg, na.rm = TRUE),
    Other_veg_median = median(Other_veg, na.rm = TRUE),
    Other_veg_se = sd(Other_veg, na.rm = TRUE) / sqrt(n()),
    Other_veg_max = max(Other_veg, na.rm = TRUE),
    Field_mean = mean(Field, na.rm = TRUE),
    Field_median = median(Field, na.rm = TRUE),
    Field_se = sd(Field, na.rm = TRUE) / sqrt(n()),
    Field_max = max(Field, na.rm = TRUE),
    Veg_Ground_mean = mean(Veg_Ground, na.rm = TRUE),
    Veg_Ground_median = median(Veg_Ground, na.rm = TRUE),
    Veg_Ground_se = sd(Veg_Ground, na.rm = TRUE) / sqrt(n()),
    Veg_Ground_max = max(Veg_Ground, na.rm = TRUE),
    Tot_Ground_mean = mean(Tot_Ground, na.rm = TRUE),
    Tot_Ground_median = median(Tot_Ground, na.rm = TRUE),
    Tot_Ground_se = sd(Tot_Ground, na.rm = TRUE) / sqrt(n()),
    Tot_Ground_max = max(Tot_Ground, na.rm = TRUE),
    Tot_Veg_mean = mean(Total_veg, na.rm = TRUE),
    Tot_Veg_median = median(Total_veg, na.rm = TRUE),
    Tot_Veg_se = sd(Total_veg, na.rm = TRUE) / sqrt(n()),
    Tot_Veg_max = max(Total_veg, na.rm = TRUE)
  )

# Pivot to long format for variables
reveg_long <- reveg_summary %>%
  pivot_longer(
    cols = -c(Treatment, t_year),
    names_to = c("Variable", "Statistic"),
     names_pattern = "(.*)_(mean|median|se|max)"
  )

# Pivot to wide format for years, ensuring correct order
reveg_wide <- reveg_long %>%
  pivot_wider(
    names_from = c(t_year, Statistic),
    values_from = value,
    names_glue = "{t_year}_{Statistic}"
  )


# Arrange the columns to have Variable, Treatment first, then statistics for each year
ordered_columns <- c(
  "Variable", "Treatment",
  "0_mean", "0_median", "0_se", "0_max",
  "2_mean", "2_median", "2_se", "2_max",
  "3_mean", "3_median", "3_se", "3_max"
)

reveg_wide <- reveg_wide %>%
  select(any_of(ordered_columns))

# Order the Treatment column
treatment_order <- c("C", "M", "S", "R")

# Arrange by Variable and then by specified Treatment order
reveg_wide <- reveg_wide %>%
  arrange(Variable, factor(Treatment, levels = treatment_order))

reveg_wide
```

## Plots with M treatment with already Sphagnum from the start

```{r}
plots_with_sphagnum <- reveg_var %>%
  filter(Treatment == "M", Sphagnum > 0, t_year == 0) %>%
 select(Location, Sublocation, Block, Treatment_ID) %>%
  distinct()

# Filter the whole dataset by joining with the specific combinations
reveg_var_MSph <- reveg_var %>%
  inner_join(plots_with_sphagnum, by = c("Location", "Sublocation", "Block", "Treatment_ID")) %>%
  filter(Treatment == "M")

```

```{r}
plot_SP_MSph <- reveg_var_MSph %>%
  filter(!is.na(Sphagnum)) %>%  # Exclude NA values in the Sphagnum column
  mutate(Treatment = factor(Treatment, levels = c("C", "M", "S", "R"))) %>% # Rearrange Treatment levels
  ggplot(aes(x = Treatment, y = Sphagnum, color = as.factor(t_year), fill=as.factor(t_year))) +   
  geom_violin(scale = "width", width= 0.8, alpha=0.7) +   
  geom_point(pch = 21, size=1, position = position_jitterdodge()) +
  ylab("Sphagnum cover") +
  labs(fill= "Growth seasons since treatment") +
  theme_minimal() +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2") +
  guides(color = "none")

plot_SP_MSph
```

```{r}
# Function to calculate standard error
se <- function(x) {
  sd(x) / sqrt(length(x))
}

# Filter for t_year == 3 and calculate summary statistics
summary_stats <- reveg_var_MSph %>%
  filter(t_year == 3) %>%
  summarise(
    mean_Sphagnum = mean(Sphagnum, na.rm = TRUE),
    median_Sphagnum = median(Sphagnum, na.rm = TRUE),
    se_Sphagnum = se(Sphagnum),
    max_Sphagnum = max(Sphagnum, na.rm = TRUE)
  )

# View the summary statistics
summary_stats
```
