---
title: "2.2 Hypothesis testing Sphagnum cover"
author: "MarteF"
format: html
editor: visual
---

# Packages needed

```{r}
library(brms) # bayesian approach to glmm, but with easier steps 
library(posterior) # tools for working with posterior and prior distributions 
library(lme4) # fit GLMM in frequentist framework 
library(tidyverse) # everything for data wrangling 
library(dplyr) # some more for data wrangling 
library(ggplot2) # everything for figures 
library(rstan) 
library(bayesplot) 
library(loo)  
library(sjPlot)
library(insight)
rstan_options(auto_write = TRUE) 
options(mc.cores = parallel::detectCores()) #To Make Stan run faster?
```

# Upload data

```{r}
reveg_var <- read_delim('../data/reveg_var.csv', 
                                  delim = ',',
                                  col_names = TRUE)

#Fill in values of Slope and Roughness from the initial year unto the following years:
reveg_var <- reveg_var %>%
  group_by(Location, Sublocation, Block, Treatment_ID) %>%
  mutate(
    Roughness = if_else(is.na(Roughness), first(Roughness, order_by = t_year), Roughness),
    Slope = if_else(is.na(Slope), first(Slope, order_by = t_year), Slope)
  ) %>%
  ungroup()
```

```{r}
reveg_var <- reveg_var %>% 
  select( -1) %>%
  mutate_at(c('year', 'month', 'Location', 'Sublocation', 'Block', 'Treatment_ID', 'Treatment', 'Name', 'Station_ID'), as.factor) %>%
  mutate_at(c('Pb', 'Al', 'B', 'Zn', 'Cu', 'Mn', 'Fe', 'P'), as.numeric) %>%
  select('Date','year','month', 'Name','Location','Sublocation', 'Block', 'Treatment', 'Treatment_ID', 'Station_ID', everything()) %>%
  rename(WT_consecutive_days_below = consecutive_days_count, WT_total_days_below=total_days, WT_mean=gs_mean, WT_max=gs_max, WT_min=gs_min) %>%
  rename(NO3_N = 'NO3-N',NH4_N = 'NH4-N')
  
```

# Creating new dataset for year 3 only

```{r}
H1.Sph <- reveg_var %>%
  filter(t_year==3) %>%
  filter(Treatment!='R') %>%
  select(Location, Sublocation, Block, Treatment, Roughness, Slope, Sphagnum, Change_Total, Sphagnum_nozero, Sphagnum_int, Sphagnum_no1, Sphagnum_beta)



```

```{r}
WT <- reveg_var %>%
  filter(Treatment != 'R') %>%
  filter(!is.na(Station_ID)) %>%
  select(c(year, Location, Sublocation, Station_ID, t_year, WT_consecutive_days_below, WT_mean, WT_max, WT_min, WT_total_days_below)) %>%
   distinct(year, Station_ID, .keep_all = TRUE)

WT2 <- WT %>%
  group_by(Location, Sublocation) %>%
 summarise(WT_consecutive_days_below_max = max(WT_consecutive_days_below),
           WT_mean = mean(WT_mean, na.rm = TRUE),
           WT_total_days_below_sum = sum(WT_total_days_below))

WT2
```

```{r}
H1.Sph <- full_join(H1.Sph, WT2, by=c('Location', 'Sublocation'))
rm(WT, WT2)

H1.Sph <- H1.Sph %>%
  filter(Treatment!='R')
```

# Hypothesis testing

These are my three hypothesis:

Sphagnum cover would

H1) respond positively to *Sphagnum* reintroduction (with mulch), *OR* that  

H2) Sphagnum cover was indifferent to the addition of *Sphagnum* fragments, but reacted to the added mulch, OR,  

H3) Sphagnum cover would be better explained by a model that only includes water table parameters. 

H2 represents the most complex model, and this was fitted like this: 

*Sphagnum cover \~ Treatment + mean_seasonal_WT + consecutive_day_with_WT_below\_-0.2m_within_season + (1\|Location/Sublocation)* 

```{r}

# H1.Sph.beta<-   brm(Sphagnum_beta/100 ~ Treatment + WT_total_days_below_sum + WT_consecutive_days_below_max + WT_mean + (1 | Location/Sublocation),
#                    data=H1.Sph,
#                  Beta(link = "logit", link_phi = "log"),
#                 chains = 4, # nb of chains
#                  iter = 4000, # nb of iterations, including burnin
#                  warmup = 1000, # burnin
#                  thin = 3,
#                 control = list(adapt_delta = 0.99, max_treedepth = 15))

#saveRDS(H1.Sph.beta, 'RDS-files/H1.Sph.beta.RDS')
readRDS(H1.Sph.beta, 'RDS-files/H1.Sph.beta.RDS')
```

```{r}
summary(H1.Sph.beta)
```

```{r}
plot(H1.Sph.beta)
```

```{r}
# Bayesian R-squared
r2 <- bayes_R2(H1.Sph.beta)
print(r2)


```

```{r}
#Generate interaction model
init_values <- function() {
  list(
    b = rep(0, 6),  # assuming 6 fixed effect coefficients
    sd_1 = 0.1, 
    sd_2 = 0.1, 
    z_1 = rnorm(length(unique(H1.Sph$Location)), 0, 0.1),
    z_2 = rnorm(length(unique(H1.Sph$Sublocation)), 0, 0.1)
  )
}

H2.Sph.beta <- brm(
  Sphagnum_beta/100 ~ Treatment*(WT_total_days_below_sum + WT_consecutive_days_below_max + WT_mean) + (1 | Location/Sublocation),
  data = H1.Sph,
  family = Beta(link = "logit", link_phi = "log"),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  thin = 3,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  init = init_values
)

#saveRDS(H2.Sph.beta, 'RDS-files/H2.Sph.beta.RDS')
#readRDS(H2.Sph.beta, 'RDS-files/H2.Sph.beta.RDS')
```

```{r}
H2.Sph.beta <- brm(
  Sphagnum_beta/100 ~ Treatment*(WT_total_days_below_sum) + (1 | Location/Sublocation),
  data = H1.Sph,
  family = Beta(link = "logit", link_phi = "log"),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  thin = 3,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)
```

```{r}
init_values <- function() {
  list(
    b = rep(0, 6),  # Assuming 6 fixed effect coefficients; adjust as necessary
    sd_1 = 0.1, 
    sd_2 = 0.1, 
    z_1 = rnorm(length(unique(H1.Sph$Location)), 0, 0.1),
    z_2 = rnorm(length(unique(H1.Sph$Sublocation)), 0, 0.1),
    shape1 = 1,  # Initial value for the first shape parameter
    shape2 = 1   # Initial value for the second shape parameter
  )
}

H2.Sph.beta <- brm(
  Sphagnum_beta/100 ~ Treatment*(WT_total_days_below_sum + WT_consecutive_days_below_max + WT_mean) + (1 | Location/Sublocation),
  data = H1.Sph,
  family = Beta(link = "logit", link_phi = "log"),
  chains = 4,
  iter = 4000,
  warmup = 1000,
  thin = 3,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  init = init_values
)
```

```{r}
summary(H2.Sph.beta)
```

```{r}
plot(H2.Sph.beta)
```

```{r}
# Bayesian R-squared
r2 <- bayes_R2(H2.Sph.beta)
print(r2)
```

```{r}
#Generate base model
H3.Sph.beta <- brm(Sphagnum_beta/100 ~ WT_total_days_below_sum + WT_consecutive_days_below_max + WT_mean + (1 | Location/Sublocation),
                   data=H1.Sph,
                 Beta(link = "logit", link_phi = "log"),
                chains = 4, # nb of chains
                 iter = 4000, # nb of iterations, including burnin
                 warmup = 1000, # burnin
                 thin = 3,
                control = list(adapt_delta = 0.99, max_treedepth = 15))

#saveRDS(H3.Sph.beta, 'RDS-files/H3.Sph.beta.RDS')
readRDS(H3.Sph.beta, 'RDS-files/H3.Sph.beta.RDS')
```

```{r}
summary(H3.Sph.beta)
```

```{r}
# Bayesian R-squared
r2 <- bayes_R2(H3.Sph.beta)
print(r2)
```

```{r}
# Generate predictions from the models
pred_H1 <- posterior_predict(H1.Sph.beta)
pred_H3 <- posterior_predict(H3.Sph.beta)


#Compare variance explained by full model vs submodel
var_H1<- apply(pred_H1, 2, var)
var_H3 <- apply(pred_H3, 2, var)

```

```{r}
# Calculate variance explained
prop_var_Treatment <- 1 - var_H3 / var_H1

mean(prop_var_Treatment)
cat("Proportion of variance explained by Treatment: ", mean(prop_var_Treatment), "\n")

```

```{r}
conditional_effects(H1.Sph.beta)
```

```{r}
loo_H1 <- loo(H1.Sph.beta)
loo_H3 <- loo(H3.Sph.beta)

loo_compare(loo_H1, loo_H3)
```

```{r}
pp_check(H1.Sph.beta, ndraws = 100, type = 'dens_overlay')
```

```{r}
pp_check(H3.Sph.beta, ndraws = 100, type = 'dens_overlay')
```
